ABM simulating short-term mating behaviors

### 2022-05-27 submission/manuscript.tex
minor language edits to the manuscript

\textsuperscript{2} xxxx

@@

Gender differences in short-term mating behaviors have been a
well-documented phenomenon in human sexuality research. Existing studies
usually conflate gender differences in mating preferences with
differences in sexual behaviors, which is theoretically dubious. Using
an agent-based model, the present study investigated the circumstances
under which men and women's differential preferences for short-term
mating would result in gender differences in short-term mating
behaviors. The results from the model suggested that heterosexual men
and women had the same average number of short-term mating experiences
and short-term mates even when men had stronger preferences for
short-term mating. Gender differences in mating behaviors only emerged
when heterosexual men and women in the mating pool (i.e., those with a
non-zero number of short-term mating experiences) were considered.
Moreover, gay men had a higher average number of both outcomes than
lesbian women as well as than heterosexual men when men had stronger
preferences for short-term mating. These results suggest that
theoretically speaking, even when gender differences in mating
preferences existed, those in short-term mating behaviors only occurred
among particular populations, or when men's preferences for short-term
mating are not constrained by those of women. Suggestions for future
research in human mating psychology and behaviors were provided.

~

@@

Gender differences in sexual behaviors, especially short-term mating
behaviors, have been a well-documented phenomenon in human sexuality
research. For example, research has found that heterosexual men, as
compared to heterosexual women, have a higher average number of past
short-term sexual partners (Oliver \& Hyde, 1993; Petersen \& Hyde,
2010; Rissel et al., 2014), have short-term mating more frequently
(Petersen \& Hyde, 2010), and engage in extramarital sex more often
(Petersen \& Hyde, 2010). These gender differences have also been
observed between gay men and lesbian women (Bailey et al., 1994; Bryant
\& Demian, 1994; Peplau et al., 1997, 2004).

In the existing literature, gender differences in sexual behaviors are
not usually distinguished from those in attitudes towards or preferences
for short-term mating. Some researchers study gender differences on the
two levels simultaneously without a conceptual distinction (e.g.,
Petersen \& Hyde, 2010), whereas some implicitly equate the two,
assuming that behavioral differences are a direct expression of the
psychological ones (e.g., Schmitt et al., 2001).

@@

The present study took gender differences in mating preferences as a
presumption, which was conceptualized by sexual strategy theory (Buss \&
Schmitt, 1993) and repeatedly found support by empirical studies (e.g.,
Schmitt, 2003; Walter et al., 2020). Using an agent-based model, we
investigated whether men and women's differential preferences for
short-term mating resulted in any gender differences in short-term
mating behaviors, specifically, in the number of short-term mating
experiences and short-term mates, and if they did, under what
circumstances.

@@

Gender differences in mating preferences have been studied in the light
of sexual strategy theory (Buss \& Schmitt, 1993). It posits that since
the minimum investment that men devote to their offspring (contribution
of sperm through one sexual act) is lower than that of women (gestation,
labor, and lactation), men tend to have short-term mating as a larger
part of their mating strategy than women do. This is because women's
high minimum investment leads to higher opportunity cost (vs.~men) if
they have sex with a suboptimal partner and give birth to offspring with
a low survival chance. As a result, men may have a greater interest in
short-term mating, desire a larger number of short-term mates in a given
period, and be less selective with respect to accepting a potential mate
(Buss \& Schmitt, 1993).

The hypotheses derived from sexual strategy theory have received
extensive support from empirical studies. For example, men reported to
be currently seeking short-term mates to a larger extent than women did
(Buss \& Schmitt, 1993; Schmitt et al., 2001; Schmitt, 2003), and a
larger proportion of them were in any way seeking short-term mates
(vs.~not seeking) (Schmitt, 2003). These have been found among U.S.
college students (Buss \& Schmitt, 1993; Schmitt et al., 2001) as well
as in ten world regions in a cross-cultural study (Schmitt, 2003).
Similarly, men across the world also reported to desire more short-term
sexual partners compared to women for a number of given time periods
(e.g., a month, a year) (Buss \& Schmitt, 1993; McBurney et al., 2005;
Schmitt, 2003). The gender differences were significant regardless of
whether they were estimated by mean (Buss \& Schmitt, 1993; Schmitt et
al., 2001; Schmitt, 2003), median (Schmitt et al., 2001; Schmitt, 2003),
or percentage (Schmitt, 2003) statistics.

As for mating standards, men are less selective in accepting someone as
a potential short-term mate. For example, studies using U.S. college
samples found that the minimum percentile ranks that men found
acceptable for a potential short-term sexual partner were lower than
those accepted by women, both in terms of overall desirability and
individual traits (e.g., social status, attractiveness) (Kenrick et al.,
1990; Regan, 1998). A study also found that when presented with
identical descriptions of potential mates, men on average rated them as
more desirable than women did (Wiederman \& Dubois, 1998).

Some evidence shows that gender differences in mating preferences also
existed between gay men and lesbian women. A survey study using a
community sample from the U.S. found that gay men were more interested
in short-term mating than lesbian women were, and that this gender
difference was comparable to that existing among heterosexual
individuals (Bailey et al., 1994). An indirect piece of evidence on
differential standards for short-term mates comes from a recent study
finding that significantly more gay men than lesbian women reported to
have accepted a casual sexual offer from a same-gender person (Matsick
et al., 2021). Past studies suggest that the gender difference in the
acceptance rate of casual sexual offers may originate from men and
women's differential standards for short-term mates (Conley et al.,
2011; Hald \& Høgh-Olesen, 2010). Thus, as a postulation, gay men may
also have lower standards than lesbian women for short-term mates.

@@

Since heterosexual sex involves both a man and a woman in most cases,
men's interest in short-term mating can be constrained by women's
preference for long-term relationships (Archer, 2019; Symons, 1979).
This is because men's short-term mating preferences can only translate
into behaviors when there are women willing to have sex with them. When
a new short-term mating encounter occurs, it counts towards men's total
number of short-term mating encounters as well as towards women's.
Similarly, a new short-term mate counts towards both men and women's
total number of short-term mates. Therefore, the total number of
short-term mating experiences and short-term mates must be equal between
heterosexual men and women on a population level. However, it is
important to note that although heterosexual men's total number of
short-term mating is equal to that of women, this does not necessarily
mean that the total number of men and women who have ever had short-term
mating must be equal. It is possible that there is a smaller proportion
of men (vs.~the proportion of women) who contribute to the total number
of short-term mating. Following this line of reasoning, we would expect
to observe no gender difference in short-term mating behaviors among
heterosexual individuals even when there were gender differences in
mating preferences.

As a comparison, in the cases of gay men and lesbian women, men's
preferences are not constrained by women's, but only by those of other
men, who, arguably, have more similar preferences. This would allow for
a more direct behavioral expression of men's mating preferences. The
notion that gay men have less restricted preferences was borne out in
the proportion of individuals who have engaged in extradyadic sex
(Peplau et al., 2004). A study showed that the proportions of
heterosexual men and women who have engaged in extradyadic sex were 26\%
and 21\%, respectively, while the statistics among gay men and lesbian
women were 82\% and 28\%, respectively (Peplau et al., 2004). Therefore,
we would theoretically expect to observe gender differences in
short-term mating behaviors between gay men and lesbian women. Moreover,
we would also expect to find that gay men, as compared to heterosexual
men, engaged in more short-term mating behaviors due to the lessened
constraint on their mating preferences.

@@

By using a spatial agent-based model, the present study investigated
whether men and women's different preferences for short-term mating
resulted in any gender differences in short-term mating behaviors, and
if they did, under what circumstances. Preferences for short-term mating
were operationally defined as the interest in short-term mating and
standards for short-term mates (Buss \& Schmitt, 1993). Short-term
mating behaviors were operationally defined as the number of short-term
mating experiences and the number of past short-term mates (Petersen \&
Hyde, 2010).

We modeled this process among both heterosexual individuals and gay men
and lesbian women to examine whether constraints set by the opposite
sex's preferences would affect short-term mating behaviors. Individuals'
sexual orientation was conceptualized in terms of behaviors only in our
model. Heterosexual men had short-term mating or formed a long-term
relationship only with women, while gay men only with other men, and
vice versa for women.

We raised the following hypotheses in the present study: when there were
gender differences in preferences for short-term mating, 1) there would
be gender differences in short-term mating behaviors among gay men and
lesbian women, with gay men being engaged in more such behaviors, and 2)
gay men would engage in more short-term mating behaviors as compared to
heterosexual men.

@@

We developed an agent-based model to represent a simplified environment
where men and women can move around, search for mates, and form
long-term and/or short-term relationships. Space and time are modeled as
discrete variables. Space was represented as discrete locations on a
two-dimensional 33*33 lattice. Agents' movement in the space was not
meant to simulate physical movement but a state of encountering
potential mates. Staying at one location represented being committed to
a long-term relationship.

The model measured two outcomes: (1) the number of short-term mating
experiences of men and women; (2) the number of short-term mates of men
and women. Additionally, we also measured the number of men and women in
the mating pool (i.e., those with a non-zero number of short-term mating
experiences). The average numbers of short-term mating experiences and
short-term mates were calculated by taking the average among the whole
population of men/women (\emph{N\textsubscript{m}} =
\emph{N\textsubscript{w}} = 150) and among those who had engaged in
short-term mating.

@@

The interest in short-term mating was modeled as an agent's likelihood
of deciding to have short-term mating at each time step. The standards
for short-term mates were modeled as the minimum desirability of a
potential mate with whom an agent was willing to have sex. The values of
the two parameters were set based on empirical findings in human mating
psychology. When there was a gender difference in short-term mating
likelihood, men had a 40\% likelihood of deciding to engage in
short-term mating, while women had a likelihood of 25\% (Buss \&
Schmitt, 1993; Schmitt et al., 2001; Schmitt, 2003). When there was no
gender difference in short-term mating likelihood, both women and men
had a likelihood of 25\%. When there was a gender difference in mating
standards, men had a mating standard of 3 (the highest possible value =
10), while women had a standard of 5 (Kenrick et al., 1990; Regan,
1998). When there was no gender difference in mating standards, both men
and women had a standard of 5.

@@

\hypertarget{heterosexual-procedures.}{%
\paragraph{Heterosexual procedures.}\label{heterosexual-procedures.}}

At each time step, agents first checked whether they were in a long-term
relationship. If they were not in a long-term relationship, they set
their heading randomly (if they were within the movement range from the
birthplace) or faced the birthplace (if they were out of the movement
range from the birthplace) and moved by a random distance. The random
distance was less than half of the movement range. If they were in a
long-term relationship, they did not move. Then, the agents decided on
whether to engage in short-term mating by a chance of the chosen
likelihood (either 25\% or 40\%).

Women checked to see if any men were at the same location. If there
were, they randomly chose one of them as their potential short-term
mate. If both men and women met each other's mating standards and both
of them decided to engage in short-term mating, they had sex. This would
result in increasing both parties' number of short-term mating
experiences by one. They also recorded each other on their lists of past
short-term mates if they were not on the lists yet.

Then, women randomly selected one man at the same location as their
potential long-term partner. There was a 10\% of chance that a pair
would form a long-term relationship. After forming a long-term
relationship, both men and women changed to coupled status and register
each other as the long-term partner.

\hypertarget{gay-men-and-lesbian-women-procedures.}{%
\paragraph{Gay men and lesbian women
procedures.}\label{gay-men-and-lesbian-women-procedures.}}

At each time step, men and women moved and decided on short-term mating
as the agents did in the heterosexual procedures. Half of the agents (75
men and 75 women) were also given the initiator status. The initiators
checked to see whether there were other agents at the same location. The
rest of the procedures were identical to the heterosexual procedures,
except that the agents only chose those with the same gender as
potential short-term mates or long-term partners.

@@

\hypertarget{statistical-analysis}{%
\subsection{Statistical analysis}\label{statistical-analysis}}

Statistical analyses were conducted using R version 4.1.3 (R Core Team,
2022) and the figures were generated by the ggplot2 package (Wickham,
2016). Two-tailed independent samples \emph{t}-tests were used for all
statistical comparisons. The data were assumed to be normally
distributed within each condition but was not formally tested.

@@

The results from Experiment 1 showed that when there were gender
differences in preferences for short-term mating, there was no gender
difference in the average number of short-term mating experiences among
heterosexual individuals (\emph{M\textsubscript{m}} = 1.43,
\emph{SD\textsubscript{m}} = 0.24; \emph{M\textsubscript{w}} = 1.43,
\emph{SD\textsubscript{w}} = 0.24), Cohen's \emph{d} = 0. Nor was there
a difference in the average number of short-term mates
(\emph{M\textsubscript{m}} = 0.80, \emph{SD\textsubscript{m}} = 0.11;
\emph{M\textsubscript{w}} = 0.80, \emph{SD\textsubscript{w}} = 0.11),
Cohen's \emph{d} = 0 (Figure 1).

\begin{figure}[h]
  \centering
  \includegraphics[width=100mm]{figures/fig1_men_vs_women.pdf}
  \caption{\textbf{Short-term mating behaviors of men and women after 1,000 time steps in the model when gender differences existed in mating preferences.} Violin plots summarizing the outcome variables separately for heterosexual individuals and gay men and lesbian women. Plot (A) shows the average number of short-term mating experiences, and plot (B) shows the average number of short-term mates. Central points show mean values and whiskers represent standard errors, but the standard errors are small and overlap to form a single bar. Statistics were calculated using the full population of men and women in the model (\textit{N\textsubscript{m}} = \textit{N\textsubscript{w}} = 150).}
\end{figure}

@@

In Experiment 1, however, when the means of the outcome variables were
calculated among men and women in the mating pool (the number of
individuals in the mating pool: \emph{M\textsubscript{m}} = 56.85,
\emph{SD\textsubscript{m}} = 5.99; \emph{M\textsubscript{w}} = 77.04,
\emph{SD\textsubscript{w}} = 7.53), there were gender differences in
short-term mating behaviors when men and women had differential
preferences for short-term mating. The average number of short-term
mating experiences was higher among heterosexual men (\emph{M} = 3.78,
\emph{SD} = 0.50) than among heterosexual women (\emph{M} = 2.79,
\emph{SD} = 0.35), \emph{t}(4,998) = -81.53, \emph{p} \textless{} .001,
Cohen's \emph{d} = 2.31. Likewise, the average number of short-term
mates was higher among heterosexual men (\emph{M} = 2.12, \emph{SD} =
0.17) than among heterosexual women (\emph{M} = 1.56, \emph{SD} = 0.11),
\emph{t}(4,998) = -135.85, \emph{p} \textless{} .001, Cohen's \emph{d} =
3.84.

@@

The results comparing across the two experiments revealed that gay men
engaged in short-term mating behaviors more than heterosexual men did
(Figure 2). The average number of short-term mating experiences was
higher among gay men (\emph{M} = 2.63, \emph{SD} = 0.37) than among
heterosexual men (\emph{M} = 1.43, \emph{SD} = 0.24), \emph{t}(4,998) =
-134.87, \emph{p} \textless{} .001, Cohen's \emph{d} = 3.81. Similarly,
the average number of short-term mates was higher among gay men
(\emph{M} = 1.72, \emph{SD} = 0.19) than among heterosexual men
(\emph{M} = 0.80, \emph{SD} = 0.11), \emph{t}(4,998) = -206.91, \emph{p}
\textless{} .001, Cohen's \emph{d} = 5.85.

\begin{figure}[h]
  \centering
  \includegraphics[width=0.8\columnwidth]{figures/fig2_hetero_vs_gay_men.pdf}
  \caption{\textbf{Short-term mating behaviors of heterosexual and gay men after 1,000 time steps in the model when gender differences existed in mating preferences.} Violin plots summarizing the outcome variables. Plot (A) shows the average number of short-term mating experiences, and plot (B) shows the average number of short-term mates. Central points show mean values and whiskers represent standard errors, but the standard errors are small and overlap to form a single bar. Statistics were calculated using the full population of heterosexual and gay men in the model (\textit{N\textsubscript{hm}} = \textit{N\textsubscript{gm}} = 150).}
\end{figure}

@@

In an exploratory manner, we also compared across conditions to see
which of the two dimensions of mating preferences contributed to gender
differences in short-term mating behaviors.

Among heterosexual individuals, gender differences in short-term mating
behaviors, as calculated among individuals in the mating pool, emerged
when men and women had different standards for short-term mates.
Heterosexual men (vs.~heterosexual women) in the mating pool had a
higher average number of short-term mating experiences and short-term
mates as long as they had lower mating standards, even when no gender
difference existed in short-term mating likelihood (experiences:
\emph{M\textsubscript{m}} = 2.82, \emph{SD\textsubscript{m}} = 0.36,
\emph{M\textsubscript{w}} = 2.21, \emph{SD\textsubscript{w}} = 0.28,
Cohen's \emph{d} = 1.89; partners: \emph{M\textsubscript{m}} = 1.76,
\emph{SD\textsubscript{m}} = 0.15; \emph{M\textsubscript{w}} = 1.38,
\emph{SD\textsubscript{w}} = 0.09, Cohen's \emph{d} = 3.13). In
comparison, virtually no gender differences in short-term mating
behaviors existed when men and women had the same mating standards
(Table 1).

@@

The present study aimed at using agent-based modeling to investigate
whether men and women's different preferences for short-term mating
resulted in any gender differences in short-term mating behaviors,
specifically, in the number of short-term mating experiences and
short-term mates, and if they did, under what circumstances. We raised
two hypotheses: 1) gay men would engage in more short-term mating
behaviors as compared to lesbian women, and 2) gay men would engage in
more short-term mating behaviors as compared to heterosexual men. The
results from 1,000 time steps in a model simulating men and women's
mating behaviors supported our hypotheses. First of all, as compared to
lesbian women, gay men had higher average numbers of short-term mating
experiences and short-term mates. Secondly, gay men also had higher
average numbers of short-term mating experiences and short-term mates as
compared to heterosexual men. In contrast, we found no gender
differences in short-term mating behaviors between heterosexual men and
women, although heterosexual men in the mating pool did have higher
average numbers of short-term mating experiences and short-term mates.

>>>>

\textsuperscript{2} Faculty of Arts, Psychology and Theology, Åbo
Akademi University

@@

Gender differences in short-term mating behaviors are well-documented in
human sexuality research. Existing studies usually conflate gender
differences in mating preferences with differences in sexual behaviors,
which is theoretically problematic. Using an agent-based model, we
investigated the circumstances under which heterosexual and homosexual
men and women's differential preferences for short-term mating would
result in gender differences in short-term mating behaviors. The model
showed that when all individuals in a closed heterosexual population
were considered, men and women had the same average number of short-term
mating experiences and short-term mates even when men had stronger
preferences for short-term mating. Men (vs.~women) had a higher average
number of both experiences and mates only when heterosexual men and
women in the mating pool (i.e., those with a non-zero number of
short-term mating experiences) were considered. Moreover, when men had
stronger preferences for short-term mating than women, gay men had a
higher average number of both experiences and mates compared to both
lesbian women and heterosexual men. These results suggest that
theoretically speaking, even when gender differences in mating
preferences existed, those in short-term mating behaviors only occurred
among particular populations, or when men's preferences for short-term
mating are not constrained by those of women. Suggestions for future
research in human mating psychology and behaviors were provided.

\hfill\break

@@

Gender differences in sexual behaviors, especially short-term mating
behaviors, are well-documented in human sexuality research. For example,
research has found that heterosexual men, as compared to heterosexual
women, have a higher average number of past short-term sexual partners
(Oliver \& Hyde, 1993; Petersen \& Hyde, 2010; Rissel et al., 2014), and
engage in both short-term mating and extramarital sex more often
(Petersen \& Hyde, 2010). These gender differences have also been
observed between gay men and lesbian women (Bailey et al., 1994; Bryant
\& Demian, 1994; Peplau et al., 1997, 2004).

In the existing literature, gender differences in actual sexual
behaviors are not usually distinguished from those in attitudes towards
or preferences for short-term mating. Some researchers study gender
differences at the two levels simultaneously without a conceptual
distinction (e.g., Petersen \& Hyde, 2010), whereas some implicitly
equate the two, assuming that behavioral differences are a direct
expression of psychological ones (e.g., Schmitt et al., 2001).

@@

As a starting point, we took gender differences in mating preferences as
conceptualized by the sexual strategy theory (Buss \& Schmitt, 1993) and
repeatedly supported by empirical investigations (e.g., Schmitt, 2003;
Walter et al., 2020). Using an agent-based model, we investigated
whether men and women's differential preferences for short-term mating
would result in gender differences in short-term mating behaviors, and
if so, then under what conditions they would result in differences in
the number of short-term mating experiences and short-term mates.

@@

Gender differences in mating preferences have been studied in the light
of sexual strategy theory (Buss \& Schmitt, 1993). It posits that since
the minimum obligatory investment that men must devote to their
offspring (contribution of sperm through one sexual act) is lower than
that of women (gestation, labor, and lactation), men tend to be
relatively more interested in short-term mating compared to women. This
is because the number of offspring women can produce is limited and
cannot be increased by mating with a large number of men, whereas the
reverse is the case for men. As a result, men are predicted to not only
have a greater interest in short-term mating, but also desire a larger
number of short-term mates in a given period and be less selective with
respect to accepting a potential mate (Buss \& Schmitt, 1993).

The hypotheses derived from sexual strategy theory have received
extensive support in empirical studies. For example, men have reported
to be currently seeking short-term mates to a larger extent than women
(Buss \& Schmitt, 1993; Schmitt et al., 2001; Schmitt, 2003), and a
larger proportion of men were in any way seeking short-term mates
(vs.~not seeking) (Schmitt, 2003). Similar results have been found among
both U.S. college students (Buss \& Schmitt, 1993; Schmitt et al., 2001)
as well as in ten world regions in a cross-cultural study (Schmitt,
2003). Similarly, men across the world have also reported to desire more
short-term sexual partners compared to women over different future time
periods (e.g., a month, a year) (Buss \& Schmitt, 1993; McBurney et al.,
2005; Schmitt, 2003). The gender differences have been found regardless
of whether they were estimated using mean (Buss \& Schmitt, 1993;
Schmitt et al., 2001; Schmitt, 2003), median (Schmitt et al., 2001;
Schmitt, 2003), or percentage (Schmitt, 2003) statistics.

As for mating standards, men are less selective in accepting someone as
a potential short-term mate. For example, studies using U.S. college
samples have found that the minimum percentile ranks that men found
acceptable for a potential short-term sexual partner were lower than
those accepted by women, both in terms of overall desirability and
individual traits (e.g., social status, attractiveness) (Kenrick et al.,
1990; Regan, 1998). Another study also found that when presented with
identical descriptions of potential mates, men on average rated them as
more desirable than women did (Wiederman \& Dubois, 1998).

Some evidence shows that gender differences in mating preferences also
existed between gay men and lesbian women. A survey study using a
community sample from the U.S. found that gay men were more interested
in short-term mating than lesbian women were, and that this gender
difference was comparable to that existing among heterosexual
individuals (Bailey et al., 1994). An additional indirect piece of
evidence on differential standards for short-term mates comes from a
recent study finding that significantly more gay men than lesbian women
reported to have accepted a casual sexual offer from a same-gender
person (Matsick et al., 2021). Past studies suggest that the gender
difference in the acceptance rate of casual sexual offers may originate
from men and women's differential standards for short-term mates (Conley
et al., 2011; Hald \& Høgh-Olesen, 2010). Thus, as a postulation, gay
men may also have lower standards than lesbian women for short-term
mates.

@@

Since heterosexual sex in most cases involves one man and one woman,
men's relatively high interest in short-term mating can be constrained
by women's preference for long-term relationships (Archer, 2019; Symons,
1979). This is because men's short-term mating preferences -- with the
relatively rare exception of sexual violence -- can only translate into
behaviors when there are women willing to have sex with them. When a new
short-term mating encounter occurs, it counts both towards men's total
number of short-term mating encounters as well as towards women's.
Similarly, a new short-term mate counts towards both men and women's
total number of short-term mates. Therefore, the total number of
short-term mating experiences and short-term mates must be equal between
heterosexual men and women at the population level. However, it is
important to note that although heterosexual men's total number of
short-term mating is equal to that of women, this does not necessarily
mean that the total number of men and women who have ever had short-term
mating must be equal. It is possible that there is a smaller proportion
of men (vs.~the proportion of women) who contribute to the total number
of short-term mating. Following this line of reasoning, we would expect
to observe no gender difference in short-term mating behaviors among
heterosexual individuals even when there are gender differences in
mating preferences.

As a comparison, in the cases of gay men and lesbian women, men's
preferences are not constrained by women's, but only by those of other
men, who, arguably, have a similar high interest in short-term mating.
This would allow for a more direct behavioral expression of men's mating
preferences. The notion that gay men have less restricted preferences is
borne out in the proportion of individuals who have engaged in
extradyadic sex (Peplau et al., 2004). A study showed that the
proportions of heterosexual men and women who had engaged in extradyadic
sex were 26\% and 21\%, respectively, while the statistics among gay men
and lesbian women were 82\% and 28\%, respectively (Peplau et al.,
2004). Therefore, we would expect to observe gender differences in
short-term mating behaviors between gay men and lesbian women. Moreover,
we would also expect to find that gay men, compared to heterosexual men,
engaged in more short-term mating behaviors due to less constraints on
their mating preferences.

@@

By using a spatial agent-based model, the present study investigated
whether men and women's differential preferences for short-term mating
would result in gender differences in short-term mating behaviors, and
if they did, under what circumstances. Preferences for short-term mating
were operationally defined as the likelihood of engaging in short-term
mating when encountering a mate as well as the standards for short-term
mates (Buss \& Schmitt, 1993). Short-term mating behaviors were
operationally defined as the number of short-term mating experiences and
the number of past short-term mates (Petersen \& Hyde, 2010).

We modeled this process among both heterosexual individuals and gay men
and lesbian women to examine whether constraints set by the opposite
sex's preferences would affect short-term mating behaviors. Individuals'
sexual orientation was conceptualized in terms of behaviors only in our
model. Heterosexual men had short-term mating or formed a long-term
relationship only with women, while gay men only with other men, and
mutatis mutandis for women.

We formulated the following hypotheses in the present study: assuming
gender differences in preferences for short-term mating, 1) there would
be gender differences in short-term mating behaviors among gay men and
lesbian women, with gay men being engaged in more such behaviors, and 2)
gay men would engage in more short-term mating behaviors as compared to
heterosexual men.

@@

We developed an agent-based model to represent a simplified environment
where men and women can move around, search for mates, and form
long-term and/or short-term relationships. Space and time were modeled
as discrete variables. Space was represented as discrete locations on a
two-dimensional 33*33 lattice. Agents' movement in the space was not
meant to simulate physical movement but a state of encountering
potential mates. Staying at one location represented being committed to
a long-term relationship.

The model measured two outcomes: (1) the number of short-term mating
experiences of men and women; (2) the number of short-term mates of men
and women. Additionally, we also measured the number of men and women in
the mating pool (i.e., those with a non-zero number of short-term mating
experiences). The average numbers of short-term mating experiences and
short-term mates were calculated by taking the average among the whole
population of men and women (\emph{N\textsubscript{men}} =
\emph{N\textsubscript{women}} = 150) and among those who had engaged in
any short-term mating.

@@

The interest in short-term mating was modeled as an agent's likelihood
of deciding to have short-term mating at each time step. The standards
for short-term mates were modeled as the minimum desirability of a
potential mate with whom an agent was willing to have sex. The values of
the two parameters were set based on empirical findings in human mating
psychology. When there was a gender difference in short-term mating
likelihood, men had a 40\% likelihood of deciding to engage in
short-term mating, while women had a likelihood of 25\% (Buss \&
Schmitt, 1993; Schmitt et al., 2001; Schmitt, 2003). When there was no
gender difference in short-term mating likelihood, both women and men
had a likelihood of 25\%. When there was a gender difference in mating
standards, men had a mating standard of 3 (the highest possible value =
10), while women had a standard of 5 (Kenrick et al., 1990; Regan,
1998), that is, they would not mate with a person with a lower mate
value. When there was no gender difference in mating standards, both men
and women had a standard of 5.

@@

\hypertarget{procedures-for-heterosexual-agents.}{%
\paragraph{Procedures for heterosexual
agents.}\label{procedures-for-heterosexual-agents.}}

At each time step, agents first checked whether they were in a long-term
relationship. If they were not in a long-term relationship, they set
their heading randomly (if they were within the movement range from the
birthplace) or faced the birthplace (if they were out of the movement
range from the birthplace) and moved by a random distance. The random
distance was less than half of the movement range. If they were in a
long-term relationship, they did not move. Then, the agents decided on
whether to engage in short-term mating using the predetermined
likelihood (either 25\% or 40\%).

Women checked to see if any men were at the same location. If there
were, they randomly chose one of them as a potential short-term mate. If
both men and women met each other's mating standards and both of them
decided to engage in short-term mating, they had sex. This would result
in increasing both parties' number of short-term mating experiences by
one. They also recorded each other on their respective lists of past
short-term mates if they were not on the lists yet.

Then, women randomly selected one man at the same location as their
potential long-term partner. There was a 10\% of chance that a pair
would form a long-term relationship. After forming a long-term
relationship, both men and women changed to coupled status and
registered each other as their long-term partner.

\hypertarget{procedures-for-gay-men-and-lesbian-women-agents.}{%
\paragraph{Procedures for gay men and lesbian women
agents.}\label{procedures-for-gay-men-and-lesbian-women-agents.}}

At each time step, men and women moved and decided on short-term mating
as the agents did in the heterosexual case. Half of the agents (75 men
and 75 women) were also given an initiator status. The initiators
checked to see whether there were other agents at the same location. The
rest of the procedures were identical to the heterosexual procedures,
except that the agents only chose those with the same gender as
potential short-term mates or long-term partners.

@@

\hypertarget{statistical-analyses}{%
\subsection{Statistical analyses}\label{statistical-analyses}}

Statistical analyses were conducted using R version 4.1.3 (R Core Team,
2022) and the figures were generated by the ggplot2 package (Wickham,
2016). Two-tailed independent samples \emph{t}-tests were used for all
statistical comparisons. The data were assumed to be normally
distributed within each condition.

@@

The results from Experiment 1 showed that when there were gender
differences in preferences for short-term mating, there was no gender
difference in the average number of short-term mating experiences among
heterosexual individuals (\emph{M\textsubscript{men}} = 1.43,
\emph{SD\textsubscript{men}} = 0.24; \emph{M\textsubscript{women}} =
1.43, \emph{SD\textsubscript{women}} = 0.24), Cohen's \emph{d} = 0. Nor
was there a difference in the average number of short-term mates
(\emph{M\textsubscript{men}} = 0.80, \emph{SD\textsubscript{men}} =
0.11; \emph{M\textsubscript{w}} = 0.80, \emph{SD\textsubscript{women}} =
0.11), Cohen's \emph{d} = 0 (Figure 1).

\begin{figure}[h]
  \centering
  \includegraphics[width=100mm]{figures/fig1_men_vs_women.pdf}
  \caption{\textbf{Short-term mating behaviors of men and women after 1,000 time steps in the model when gender differences existed in mating preferences.} Violin plots summarizing the outcome variables separately for heterosexual individuals and gay men and lesbian women. Plot (A) shows the average number of short-term mating experiences, and plot (B) shows the average number of short-term mates. Central points show mean values and whiskers represent standard errors, but the standard errors are small and overlap to form a single bar. Statistics were calculated using the full population of men and women in the model (\textit{N\textsubscript{men}} = \textit{N\textsubscript{women}} = 150).}
\end{figure}

@@

In Experiment 1, however, when the means of the outcome variables were
calculated only among men and women in the mating pool (the number of
individuals in the mating pool: \emph{M\textsubscript{men}} = 56.85,
\emph{SD\textsubscript{men}} = 5.99; \emph{M\textsubscript{women}} =
77.04, \emph{SD\textsubscript{women}} = 7.53), there were gender
differences in short-term mating behaviors when men and women had
differential preferences for short-term mating. The average number of
short-term mating experiences was higher among heterosexual men
(\emph{M} = 3.78, \emph{SD} = 0.50) than among heterosexual women
(\emph{M} = 2.79, \emph{SD} = 0.35), \emph{t}(4,998) = -81.53, \emph{p}
\textless{} .001, Cohen's \emph{d} = 2.31. Likewise, the average number
of short-term mates was higher among heterosexual men (\emph{M} = 2.12,
\emph{SD} = 0.17) than among heterosexual women (\emph{M} = 1.56,
\emph{SD} = 0.11), \emph{t}(4,998) = -135.85, \emph{p} \textless{} .001,
Cohen's \emph{d} = 3.84.

@@

Gay men engaged in short-term mating behaviors more than heterosexual
men did (Figure 2). The average number of short-term mating experiences
was higher among gay men (\emph{M} = 2.63, \emph{SD} = 0.37) than among
heterosexual men (\emph{M} = 1.43, \emph{SD} = 0.24), \emph{t}(4,998) =
-134.87, \emph{p} \textless{} .001, Cohen's \emph{d} = 3.81. Similarly,
the average number of short-term mates was higher among gay men
(\emph{M} = 1.72, \emph{SD} = 0.19) than among heterosexual men
(\emph{M} = 0.80, \emph{SD} = 0.11), \emph{t}(4,998) = -206.91, \emph{p}
\textless{} .001, Cohen's \emph{d} = 5.85.

\begin{figure}[h]
  \centering
  \includegraphics[width=0.8\columnwidth]{figures/fig2_hetero_vs_gay_men.pdf}
  \caption{\textbf{Short-term mating behaviors of heterosexual and gay men after 1,000 time steps in the model when gender differences existed in mating preferences.} Violin plots summarizing the outcome variables. Plot (A) shows the average number of short-term mating experiences, and plot (B) shows the average number of short-term mates. Central points show mean values and whiskers represent standard errors, but the standard errors are small and overlap to form a single bar. Statistics were calculated using the full population of heterosexual and gay men in the model (\textit{N\textsubscript{hetero}} = \textit{N\textsubscript{gay}} = 150).}
\end{figure}

@@

We also ran analyses across conditions to see which of the two
dimensions of mating preferences contributed to gender differences in
short-term mating behaviors.

Among heterosexual individuals, gender differences in short-term mating
behaviors, when calculated among individuals in the mating pool, emerged
when men and women had different standards for short-term mates.
Heterosexual men (vs.~heterosexual women) in the mating pool had a
higher average number of short-term mating experiences and short-term
mates as long as they had lower mating standards, even when no gender
difference existed in short-term mating likelihood (experiences:
\emph{M\textsubscript{men}} = 2.82, \emph{SD\textsubscript{men}} = 0.36,
\emph{M\textsubscript{women}} = 2.21, \emph{SD\textsubscript{women}} =
0.28, Cohen's \emph{d} = 1.89; partners: \emph{M\textsubscript{men}} =
1.76, \emph{SD\textsubscript{men}} = 0.15; \emph{M\textsubscript{women}}
= 1.38, \emph{SD\textsubscript{women}} = 0.09, Cohen's \emph{d} = 3.13).
In comparison, virtually no gender differences in short-term mating
behaviors existed when men and women had the same mating standards
(Table 1).

@@

The present study used agent-based modeling to investigate whether men
and women's differential preferences for short-term mating would result
in gender differences in short-term mating behaviors, and if so, then
under what conditions they would result in differences in the number of
short-term mating experiences and short-term mates. We formulated two
hypotheses: 1) gay men would engage in more short-term mating behaviors
as compared to lesbian women, and 2) gay men would engage in more
short-term mating behaviors as compared to heterosexual men. The results
from 1,000 time steps in a model simulating men and women's mating
behaviors provided strong evidence in favor of our hypotheses. First of
all, as compared to lesbian women, gay men had higher average numbers of
short-term mating experiences and short-term mates. Secondly, gay men
also had higher average numbers of short-term mating experiences and
short-term mates as compared to heterosexual men. In contrast, we found
no gender differences in short-term mating behaviors between
heterosexual men and women, although heterosexual men in the mating pool
had higher average numbers of short-term mating experiences and
short-term mates.

As expected, heterosexual men and women did not differ in short-term
mating behaviors despite their differential preferences for short-term
mating. This was because heterosexual men and women had an equal total
number of both short-term mating experiences and short-term mates. This
is not surprising since the sex ratio was 1:1 in our model. It therefore
follows that the average number of short-term mating experiences and
short-term mates must be equal between men and women as well. However,
we did find that when we only looked at individuals in the mating pool
(from which more men than women were excluded), men engaged in more
short-term mating behaviors as compared to women. This was because there
were more women than men in the mating pool, resulting in lower averages
among heterosexual women despite the equal numbers of experiences and
mates in total.
